---
title: "Cyclistic"
author: "JQH"
date: "10/4/2021"
output: html_document
---

# Cyclistic Data Analysis Project (Track 1 of Google Data Analytics Certification)

## ASK Step

The client has provided a number of files split into quarters and contain the ride sharing data for 2021. this data will be loaded , explored and cleaned as needed to prepare for analysis.

The main team involved for this project are :

-   Lily Moreno -> director of marketing and will be responsible to form the intended new strategy based on the data insights

-   Cyclistic marketing analytics team -> the collected insights from this analysis will be combined with other work from this team for the final analysis

-   Cyclistic Exec team -> final approval on going ahead with the program will come from this team.

The main business Task required from this part of the project is :

> Understanding the main difference between the Casual and Member riders of the Cyclistic company. and based on the insights and Data , a new marketing strategy will be formed.

## Prepare Step

The data sets were downloaded from client and stored temporarily for analysis. all the data sets for 2021 were given in a csv file format and will be inspected for any issues by using R analysis tools and later after cleaning.

#### Data Ethics and Privacy

Since the data is coming directly from Motivate international and is assumed to be fit for the business case study purpose. The data is under a [licence](https://ride.divvybikes.com/data-license-agreement) agreement which prohibits storing or distributing this data in any way. Inferring or atempting to relate this data to any individuals or persons who have used the cyclistic service.

## Process Step

The process step will be to combine all the datasets into one main set and start the cleaning and analysis process. since we are using an R markdown file all cleaning steps will be documented below in R code blocks.

first we load our required libraries for data manipulation and processing

```{r Loading Libraries and Data into the workspace , message=FALSE, warning=FALSE}
# load libraries required for processing and analysis 

library(tidyverse) # used for the data wrangling and data manipulation 
library(janitor) # used for data cleaning 
library(DataExplorer)
library(lubridate) # used to easily convert data types of date and time formats 

# Clean the files environment in the R work space ( best practice to always start fresh)

rm(list=ls())

# set the current work directory to be the 2021 directory 

setwd("../Working_Dir/CSV/2021")

# create Data frames from all the files in the directory 

df1 <- read.csv("202101-divvy-tripdata.csv")
df2 <- read.csv("202102-divvy-tripdata.csv")
df3 <- read.csv("202103-divvy-tripdata.csv")
df4 <- read.csv("202104-divvy-tripdata.csv")
df5 <- read.csv("202105-divvy-tripdata.csv")
df6 <- read.csv("202107-divvy-tripdata.csv")
df7 <- read.csv("202107-divvy-tripdata.csv")
df8 <- read.csv("202108-divvy-tripdata.csv")
```

```{r examining each file to explore the data}
# examine each file to ensure consistent col names and investigate the data 

str(df1)
str(df2)
str(df3)
str(df4)
str(df5)
str(df6)
str(df7)
str(df8)



```

Looking at the Structure of each dataset we can see that they can be easily combined into one master file for easier cleanup and exploring

```{r combine all the dataframes into one for cleanup prior to any analysis  }
# Create one Variable Data frame with all the df's combined with rbind function (this requires the data frame structure to be matching)

bike_rides_2021 <- rbind(df1,df2,df3,df4,df5,df6,df7,df8)

# View a Summary of the data to see the data structure and investigate any issues with the data

summary(bike_rides_2021)

```

```{r Clean up the data and transform it for analysis }

# clean up the data by removing any empty rows, cols and create the date variable with lubridate  
bike_rides_2021 <- janitor::remove_empty(bike_rides_2021 , which = c("rows","cols"))
bike_rides_2021 <- remove_empty(bike_rides_2021)
# convert the columns to appropriate date types

bike_rides_2021$started_at <- lubridate::ymd_hms(bike_rides_2021$started_at)
bike_rides_2021$ended_at <- lubridate::ymd_hms(bike_rides_2021$ended_at)

# add the duration column to the data to know the duration of all trip

bike_rides_2021$duration <- bike_rides_2021$ended_at - bike_rides_2021$started_at
lubridate::hms(bike_rides_2021$duration
               
bike_rides_2021 <- bike_rides_2021 %>% 
  unique()

```

## Visualize and Share Step

```{r Visualize and Share the insights }
# Start to visulize the data 

library(ggplot2)

bike_rides_2021 %>%  filter(duration < 100 & start_lat > 41.9)
ggplot(data = bike_rides_2021)+
  geom_line(mapping = aes(x= bike_rides_2021$member_casual, y = bike_rides_2021$duration , color = bike_rides_2021$rideable_type))

bike_rides_2021 %>%  filter(duration > 600 & start_lat < 41.9)
ggplot(data = bike_rides_2021)+
  geom_line(mapping = aes(x= bike_rides_2021$member_casual, y = bike_rides_2021$duration , color = bike_rides_2021$rideable_type) )
  

```

## ACT and Recommend Step
